Welfare Expectations Across U.S. Economic Cycles

### 2024-04-23 paper/paper.qmd
Update figure 1

---
title: "Welfare Expectations Across U.S. Economic Cycles"
subtitle: "Analysis of the the American public's views on government provision of benefits under different economic conditions over the past 50 years"
author: 
  - Tianen (Evan) Hao
thanks: "Code and data are available at: https://github.com/ShadyEvan4830/Welfare_Expectations_and_The_Economy.git."
date: March 31, 2024
date-format: long
abstract: "A country's economic conditions often influence public perceptions of government welfare practices. However, do all citizens truly meet the general expectation of higher welfare during economic downturns and higher welfare during economic boom? Using data from the General Social Survey (GSS) Welfare, this paper examines whether the American public believes government welfare initiatives have been adequate over the past 50 years, particularly under different economic circumstances. Our analysis shows that there are widespread public calls to reduce welfare spending during good times, while in bad times, this leads to a stronger demand for government assistance. Despite this trend, journal research shows that differences in welfare expectations remain evident between different demographic groups, such as high- and low-income groups. This finding highlights the need for society to develop welfare policies for different groups and provide a basis for creating fairer and more sensitive welfare strategies amid economic fluctuations."
format: pdf
toc: true
number-sections: true
bibliography: bibliography.bib
---

@@

Public attitudes towards government welfare programs fluctuate with economic conditions, swinging from calls for austerity during prosperous times to demands for increased support in downturns. This study investigates whether there's a consistent trend of advocating for reduced welfare spending in periods of economic prosperity and whether there's a noticeable shift towards supporting more government welfare provisions during downturns, as seen during the 1982 recession, the 2008 Global Financial Crisis, and the 2020 COVID-19 pandemic. Importantly, we explore whether this trend is consistent across all public segments or if attitudes vary among different demographic groups through various journal research.

@@

We hypothesize that the public expects more benefits when the economy is low and fewer benefits when the economy is growing. The estimand is the difference in public support for increased welfare benefits if the economy were in a recession, as opposed to if the economy were experiencing growth. This is assessed through the lens of varying response groups to understand if the demand for greater welfare support is a ubiquitous response to economic hardship. Our preliminary findings reveal that public opinion is actually multifaceted. The recession significantly affected demand for increased government welfare action but with significant differences across demographic groups; different incomes or other social factors will significantly affect people's views on welfare. In addition to this we have seen a growing demand in recent years for perceived welfare needs to be enhanced.

@@

The structure of this paper starts with the current [Introduction] to explain the overall logic and content. The subsequent [Data] section outlines our methods and data-cleaning efforts. In the [Model] section, we use a linear regression model to further analyze the number of public responses to "Too Little" of welfare since 1972. The [Results] section presents findings on public welfare perceptions under different economic conditions. In [Discussion], we interpreted within the larger context of welfare policy and public expectations and are expected to trigger reflections on the study's limitations and potential directions for future research. Finally, we synthesize our main insights in [Conclusion] to highlight the urgent need for nuanced and adaptable welfare policies. [Appendix] provides further model details supporting our analysis.

@@

Survey Audience GSS focuses on adults 18 years of age and older who live in the United States and communicate in English or Spanish [@gssCodebook2022]. We worry that despite this inclusion criterion enabled a more diverse group of participants to contribute to the survey. However, those who do not meet these language requirements or are still unable to participate for health reasons are excluded from the scope of the @GSS.

@@

Screening analyses form another key component of our statistical analysis. By classifying years according to the extreme sentiment expressed about welfare spending, we identify periods particularly salient in public opinion (e.g., unusually high levels of a certain variable). This approach allows us to isolate and examine years of significant public consensus on welfare policy and compare it to years of average economic growth to understand the potential catalysts driving these sentiments.

@@

Observing the trend in the graph @fig-too-little-trend, we note an undulating pattern of "too little" responses over the fifty-year span. The trend is characterized by significant peaks and troughs, it is notable that there are periods where the counts sharply increase, suggesting spikes in the sentiment that welfare is insufficient. Following these peaks, there are also steep declines at certain intervals, indicating a drop in the number of "too little" responses. Overall, despite these fluctuations, there is a discernible upward trend in the recent 10 years (2012-2022), culminating in a prominent peak towards the end of the period under consideration. This trend suggests an increasing number of respondents perceive welfare provisions to be inadequate in more recent years.

@@

Our data analysis found trends in the “too little” category of General Social Survey responses that could indicate public perceptions of benefit adequacy. This trend corresponds to the changing economic landscape over the years, and to capture this dynamic quantitatively, we designed a linear regression model to predict the likelihood of a “too few” response depending on the year of survey. We ran the model in R [@citeR] using @rstanarm's `rstanarm` package. For comprehensive background details on model construction, variables, and diagnostics, please see the [Appendix](#sec-model-details).

@@

We posit a clear relationship between the passage of time and the rising public advocacy for more comprehensive welfare reflected in "too little" responses within the General Social Survey (GSS) data. Specifically, we hypothesize that the proportion of "Too Little" responses exhibits a consistent upward trend from 1972 to 2022, indicative of a growing public sentiment that government welfare provision is falling short of societal needs.

The linear regression model we propose serves to validate this hypothesis mathematically. If the coefficient $\beta_1$ to be positive will confirm our assumption of an upward trajectory for “too few” responses. As an illustrative example, suppose that the GSS data show a large increase in “too few” responses in the years following the implementation of the welfare reform policy, which would appear in our model as a steeper slope in these years.

@@

## Analysis of Model Results
The linear model outlined in @tbl-linear-trend and visualized in the accompanying graph suggests a positive trend in the frequency of "too little" responses over the surveyed years. This trend is evidenced by the positive coefficient for 'year' ($\beta_1$ = 3.63), which suggests that each successive year is associated with a slight increase in the proportion of "Too Little" responses.

@@

## Model Limitations
The Model section introduces a linear regression model to predict the probability of a "Too Little" response based on the survey year. While the model's simplicity is a strength, it's also a limitation. The assumption that the relationship between time and perception of welfare adequacy is linear does not account for the complexities of socioeconomic factors. For example, policy changes, like the welfare reform of the 1990s, could have nonlinear effects on public opinion that a simple linear model would not capture [@blank2005did].

The low R² value indicates that the year alone is not a strong predictor of the perception of welfare adequacy, suggesting that critical variables may be missing from the model. Other potential predictors, such as unemployment rates, economic growth indicators, or significant political events, are not included in the model but could provide additional explanatory power. For example, a year with significant job growth might see a disproportionate decrease in "Too Little" responses, which would not be accounted for by the year alone.

>>>>

---
title: "Welfare Expectations Across U.S. Economic Cycles"
subtitle: "Analysis of the the American public's views on government provision of benefits under different economic conditions over the past 50 years"
author: 
  - Tianen (Evan) Hao
thanks: "Code and data are available at: https://github.com/ShadyEvan4830/Welfare_Expectations_and_The_Economy.git."
date: March 31, 2024
date-format: long
abstract: "A country's economic conditions often influence public perceptions of government welfare practices. However, do all citizens truly meet the general expectation of higher welfare during economic downturns and lower welfare during economic boom? Using data from the General Social Survey (GSS) Welfare, this paper examines whether the American public believes government welfare initiatives have been adequate over the past 50 years, particularly under different economic circumstances. Our analysis shows that there are widespread public calls to reduce welfare spending during good times, while in bad times, this leads to a stronger demand for government assistance. Despite this trend, journal research shows that differences in welfare expectations remain evident between different demographic groups, such as high and low-income groups. This finding highlights the need for society to develop welfare policies for different groups and provide a basis for creating fairer and more sensitive welfare strategies amid economic fluctuations."
format: pdf
toc: true
number-sections: true
bibliography: bibliography.bib
---

@@

Public attitudes towards government welfare programs fluctuate with economic conditions, swinging from calls for austerity during prosperous times to demands for increased support in downturns. This study investigates whether there is a consistent trend of advocating for reduced welfare spending in periods of economic prosperity and whether there's a noticeable shift towards supporting more government welfare provisions during downturns, as seen during the 1982 recession and the 2020 COVID-19 pandemic. Importantly, we explore whether this trend is consistent across all public segments or if attitudes vary among different demographic groups through various journal research.

@@

We hypothesize that the public expects more benefits when the economy is low and fewer benefits when the economy is growing. The estimand is the difference in public support for increased welfare benefits if the economy were in a recession, as opposed to if the economy were experiencing growth. This is assessed through the lens of varying response groups to understand if the demand for greater welfare support is a ubiquitous response to economic hardship. Our preliminary findings show that public opinion is multifaceted. The recession significantly affected demand for increased government welfare action but with significant differences across demographic groups; different incomes or other social factors will significantly affect people's views on welfare. In addition to this we have seen a growing demand in recent years for perceived welfare needs to be enhanced.

@@

The structure of this paper starts with the current [Introduction] to explain the overall logic and content. The subsequent [Data] section outlines our methods and data-cleaning efforts. In the [Model] section, we use a linear regression model to further analyze the number of public responses to "too little" of welfare since 1972. The [Results] section presents findings on public welfare perceptions under different economic conditions. In [Discussion], we interpreted within the larger context of welfare policy and public expectations and are expected to trigger reflections on the study's limitations and potential directions for future research. Finally, we synthesize our main insights in [Conclusion] to highlight the urgent need for nuanced and adaptable welfare policies. [Appendix] provides further model details supporting our analysis.

@@

Survey Audience GSS focuses on adults 18 years of age and older who live in the United States and communicate in English or Spanish [@gssCodebook2022]. We worry that despite this inclusion criterion enabled a more diverse group of participants to contribute to the survey; those who do not meet these language requirements or are still unable to participate for health reasons are excluded from the scope of the @GSS.

@@

Screening analyses form another key component of our statistical analysis. By classifying years according to the extreme sentiment expressed about welfare spending, we identify periods particularly salient in public opinion (e.g. unusually high levels of a certain variable). This approach allows us to isolate and examine years of significant public consensus on welfare policy and compare it to years of average economic growth to understand the potential catalysts driving these sentiments.

@@

Observing the trend in the graph @fig-too-little-trend, we note an undulating pattern of "too little" responses over the fifty-year span. The trend is characterized by significant peaks and troughs, it is notable that there are periods where the counts sharply increase, suggesting spikes in the sentiment that welfare is insufficient. Following these peaks, there are also steep declines at certain intervals, indicating a drop in the number of "too little" responses. Overall, despite these fluctuations, there is a discernible upward trend in the recent 10 years (2012-2022), culminating in a prominent peak towards the end of the period under consideration. This trend shows an increasing number of respondents perceive welfare provisions to be inadequate in more recent years.

@@

Our data analysis found trends in the “too little” category of General Social Survey responses that could indicate public perceptions of benefit adequacy. This trend corresponds to the changing economic landscape over the years, and to capture this dynamic quantitatively, we designed a linear regression model to predict the likelihood of a “too little” response depending on the year of survey. We ran the model in R [@citeR] using @rstanarm's `rstanarm` package. For comprehensive background details on model construction, variables, and diagnostics, please see the [Appendix](#sec-model-details).

@@

We posit a clear relationship between the passage of time and the rising public advocacy for more comprehensive welfare reflected in "too little" responses within the General Social Survey (GSS) data. Specifically, we hypothesize that the proportion of "too little" responses exhibits a consistent upward trend from 1972 to 2022, indicative of a growing public sentiment that government welfare provision is falling short of societal needs.

The linear regression model we propose serves to validate this hypothesis mathematically. If the coefficient $\beta_1$ to be positive will confirm our assumption of an upward trajectory for “too little” responses. As an illustrative example, suppose that the GSS data show a large increase in “too little” responses in the years following the implementation of the welfare reform policy, which would appear in our model as a steeper slope in these years.

@@

## Analysis of Model Results
The linear model outlined in @tbl-linear-trend and visualized in the accompanying graph suggests a positive trend in the frequency of "too little" responses over the surveyed years. This trend is evidenced by the positive coefficient for 'year' ($\beta_1$ = 3.63), which suggests that each successive year is associated with a slight increase in the proportion of "too little" responses.

@@

## Model Limitations
The Model section introduces a linear regression model to predict the probability of a "too little" response based on the survey year. While the model's simplicity is a strength, it's also a limitation. The assumption that the relationship between time and perception of welfare adequacy is linear does not account for the complexities of socioeconomic factors. For example, policy changes, like the welfare reform of the 1990s, could have nonlinear effects on public opinion that a simple linear model would not capture [@blank2005did].

The low R² value indicates that the year alone is not a strong predictor of the perception of welfare adequacy, suggesting that critical variables may be missing from the model. Other potential predictors, such as unemployment rates, economic growth indicators, or significant political events, are not included in the model but could provide additional explanatory power. For example, a year with significant job growth might see a disproportionate decrease in "too little" responses, which would not be accounted for by the year alone.
